Modal operators for a constructive account of well quasi-orders

### 2024-11-30 slides-abmv2024.tex
Talk for ABMV 2024

\documentclass[11pt,utf8,notheorems,compress,t,aspectratio=169]{beamer}
\usepackage{etex}

@@

\usepackage[normalem]{ulem}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{manfnt}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{proof}
\usepackage{agda}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning,svg.path}
\hypersetup{colorlinks=true}

@@

\title{Towards multiversal modal operators for homotopy type theory}

\author{Ingo Blechschmidt}
\date{October 30th, 2023}

@@

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\BB}{\mathbb{B}}
\newcommand{\pp}{\mathbbm{p}}
\newcommand{\MM}{\mathbb{M}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\FF}{\mathbb{F}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\CC}{\mathbb{C}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\ccc}{\mathfrak{c}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Cov}{\mathrm{Cov}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\inv}{inv.\@}
\newcommand{\forces}{\vDash}

\setbeamertemplate{blocks}[rounded][shadow=false]

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\includegraphics[height=\paperheight]{multiverse-faded}\end{minipage}}
\begin{frame}[c]
  \centering
  \color{white}

  \bigskip
  \bigskip
  \bigskip
  \bigskip

  \scriptsize
  \textit{-- an invitation --}

  \setbeamercolor{block body}{bg=black!100}
  \begin{minipage}{0.49\textwidth}
    \begin{block}{}
      \centering\normalsize\color{white}
      \hil{Towards multiversal modal operators for homotopy type theory}
    \end{block}
  \end{minipage}

  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
\end{frame}}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[height=\paperheight]{sea-of-clouds-2}\end{minipage}}
\begin{frame}{How should this notion be formalized in HoTT?}
  \jnote{5-}{
    The presented proof rests on the law of excluded middle and hence cannot
    immediately be interpreted as a program for finding suitable indices~$i < j$.
    However, constructive proofs are also possible (for instance by
    induction on the value of a given term of the sequence), and furthermore
    constructive proofs can be extracted from the presented classical proof.
  }

  \jnote{7-}{
    The class of well quasiorders is stable under cartesian products, lists and
    trees, by Dickson's Lemma, Higson's Lemma and Kruskal's Theorem,
    respectively. However, in their naive formulations, these are merely
    theorems of classical mathematics. For general constructive results, the
    definition of ``well'' needs to be improved.

    Classical texts often employ needless negations in their definitions; this
    is an XXX
  }

  \vspace*{-1em}
  \[ \astikznodetransparentlycircled{xm}{7}\!,
    \quad \astikznodetransparentlycircled{x0}{4}\!,
    \quad \only<1-2>{\astikznodetransparentlycircled{t1}{3}}\only<3->{\astikznodecircled{t1}{mypurple}{3}}\!,
    \quad \only<1>{\ldots}\pause \astikznodetransparentlycircled{x1}{1}\!,
    \quad \only<2>{\ldots}\pause \astikznodecircled{t2}{mypurple}{8}\!,
    \quad \only<3>{\ldots} \visible<4->{\astikznodetransparentlycircled{x2}{2}\!,}
    \quad \only<4->{\ldots} \]
  {\centering\begin{tikzpicture}[remember picture,overlay]
    \node[draw=mypurple, circle, thick, inner sep=0.1em] (t3) {\scriptsize$\leq$};
    \path[draw=mypurple,thick]
      (t1)
      to [out=-90, in=180] (t3)
      to [out=0, in=-90] (t2);
  \end{tikzpicture}\par}
  \medskip
  \pause

  \begin{block}{}
    \justifying
    \textbf{Thm.} Every sequence~$\alpha : \NN \to \NN$ is \hil{good} in that
    there exist~$i < j$ with~$\alpha\,i \leq \alpha\,j$.
  \end{block}
  \pause
  \vspace*{-0.6em}
  \emph{Proof.} \emph{(offensive?)} By~\badbox{\textsc{lem}}, there is a
  minimum~$\alpha\,i$.
  Set~$j \defeq i + 1$. \qed\par
  \pause

  \textbf{Def.} (classically) A quasiorder~$X$ is \hil{well} iff every sequence~$\NN \to X$ is good.

  \textbf{Examples.} (classically)\ \ $(\NN,{\leq}),\ \ X \times Y,\ \ X^*,\ \ \mathrm{Tree}(X).$
  \bigskip
  \pause

  \emph{A naive formalization attempt:}

  \textbf{Def.} $\displaystyle\mathsf{Well}_\infty(X,{\leq}) \defeq
    \prod\limits_{\alpha \? \NN \to X} \, \Bigl\|\sum\limits_{i \? \NN} \sum\limits_{j \? \NN}
      i < j \times \alpha\,i \leq \alpha\,j\Bigr\|_{-1}$
  \pause
  \qquad\scalebox{1.5}{\dbend}
  \medskip

  \begin{itemize}
    \item[\bad{\xmark}] philosophically~strenuous  %: needless reference of the transfinite
    \item[\bad{\xmark}] not practical  %: cannot verify closure properties
    \item[\bad{\xmark}] not faithful(?)  %: missing out on some sequences
  \end{itemize}
\end{frame}}

\begin{frame}{Sequences provided by \textsc{lem}}
  \justifying
  \begin{block}{}
    \justifying
    \textbf{Lemma.} Let~$X$ be a well quasiorder. Let~$\alpha : \NN \to X$ be a
    sequence. Assuming~\bad{\textsc{lem}}, there merely is a monotonic
    subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
  \end{block}
  \vspace*{-0.6em}
  \emph{Proof.} The type
  \[ I \defeq \sum_{i \? \NN} \neg\sum_{j \? \NN} i < j \times \alpha\,i \leq \alpha\,j \]
  cannot be in bijection with~$\NN$, as else the~$I$-extracted subsequence
  of~$\alpha$ would not be good. By~\bad{\textsc{lem}}, the type~$I$ is finite.
  Any index~$i_0$ larger than all the indices in~$I$ is a suitable starting
  point for a monotonic subsequence. \qed
  \medskip
  \pause

  \begin{block}{}
    \justifying
    \textbf{Prop.} Let~$X$ and~$Y$ be well quasiorders.
    Assuming~\bad{\textsc{lem}},~$X \times Y$ is well.
  \end{block}
  \vspace*{-0.6em}
  \emph{Proof.} Let an infinite sequence~$\gamma : \NN \to X \times Y$ be
  given. Write~$\gamma\,k = (\alpha\,k,\beta\,k)$. By the lemma, there is a
  monotonic subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$. Because~$Y$
  is well, there are indices~$n < m$ such that~$\beta\,i_n \leq \beta\,i_m$. As
  also~$\alpha\,i_n \leq \alpha\,i_m$, the sequence~$\gamma$ is good. \qed
  \medskip
  \pause

  \dnote{\emph{We cannot trust~\bad{\textsc{lem}}-provided sequences to be
  available in the type~$\NN \to X$. Similarly with~\bad{\textsc{dc}}.}}
\end{frame}

\begin{frame}{Sequences depending on an environment}
  Let~$X$ be an hset such that there is no surjection~$\NN \twoheadrightarrow X$.
  Then the type~$\NN \to X$ misses the \hil{generic enumeration}~$\genalpha$ of~$X$.
  \medskip
  \pause

  \emph{Idea:} Approximate (fictitious) surjections~$\NN \twoheadrightarrow X$
  by finite sequences~$\sigma = x_0 \ldots x_{n-1} \? X^*$. \\
  Starting with the empty sequence~$\varepsilon$, over time, such an
  approximation can grow to
  \begin{enumerate}
    \item one of $\sigma y$, where~$y \? X$, so it becomes \emph{more defined}; or
    \item one of~$\sigma \tau$, where~$\tau \? X^*$ such that~$a \in
    \sigma\tau$, so it becomes \emph{more surjective}.
  \end{enumerate}
  \medskip
  \pause

  For~$k : \NN$ and~$a : X$, the expression~``$\genalpha\,k = a$'' does not
  denote a proposition but rather a \hil{stage-dependent proposition}~$X^* \to
  \mathsf{Prop}$, namely $\lambda x_0 \ldots x_{n-1}.\ k < n \times x_k = a$.
  \medskip
  \pause

  Given a stage-dependent proposition~$P$, $\hil{$\nabla P$}\,\sigma$ expresses that no
  matter how~$\sigma$ evolves to a better approximation~$\tau$,
  eventually~$P\,\tau$ will hold:
  \AgdaHide{
  \begin{code}
    {-# OPTIONS --no-prop --no-import-sorts --allow-unsolved-metas #-}
    open import Agda.Primitive using (Set)
    Prop = Set
    infixl 5 _∈_
    postulate
      X : Set
      X* : Set
      [] : X*
      _∷_ : X* → X → X*
      _⧺_ : X* → X* → X*
      _∈_ : X → X* → Prop
    module Foo where
  \end{code}
  }
  {\small\begin{code}
      data ∇ (P : X* → Prop) (σ : X*) : Prop where
        now : P σ → ∇ P σ
        later₁ : ((y : X) → ∇ P (σ ∷ y)) → ∇ P σ
        later₂ : (a : X) → ((τ : X*) → a ∈ σ ⧺ τ → ∇ P (σ ⧺ τ)) → ∇ P σ
  \end{code}}

  \vspace*{-0.8em}
  For instance, for~$P_a\,\sigma \defeq (a \in \sigma)$,
  we have~$\nabla P_a \, \varepsilon$. \quad ``$\genalpha$ is surjective'':
  \AgdaSymbol{(}\AgdaBound{a}\AgdaSpace{}%
  \AgdaSymbol{:}\AgdaSpace{}%
  \AgdaPostulate{X}\AgdaSymbol{)}\AgdaSpace{}%
  \AgdaSymbol{→}\AgdaSpace{}%
  \AgdaDatatype{∇}\AgdaSpace{}%
  \AgdaBound{P$_a$}\AgdaSpace{}%
  \AgdaBound{$\varepsilon$}
\end{frame}

\begin{frame}{Well quasiorders revisited}
  \emph{A naive formalization attempt:}

  \textbf{Def.} $\displaystyle\mathsf{Well}_\infty(X,{\leq}) \defeq
    \prod\limits_{\alpha \? \NN \to X} \, \Bigl\|\sum\limits_{i \? \NN} \sum\limits_{j \? \NN}
      i < j \times \alpha\,i \leq \alpha\,j\Bigr\|_{-1}$
  \pause
  \qquad\scalebox{1.0}{\dbend}
  \bigskip

  \emph{An inductive rephrasing:}

  \textbf{Def.} $\displaystyle\mathsf{Well}(X,{\leq}) \defeq
    \nabla \mathsf{Good} \, \varepsilon$, where~$\mathsf{Good}\,x_0 \ldots
    x_{n-1} \defeq \mathop{\exists}\limits_{i:\NN}\mathop{\exists}\limits_{j:\NN} (i < j \times x_i \leq
    x_j)$ and
  {\vspace*{-0.7em}\small\begin{code}
    data ∇ (P : X* → Prop) (σ : X*) : Prop where
      now : P σ → ∇ P σ
      later : ((y : X) → ∇ P (σ ∷ y)) → ∇ P σ
  \end{code}}

  \vspace*{-0.8em}
  In other words: \emph{A quasiorder~$X$ is well iff the generic sequence~$\NN \to X$
  is good.}

  \begin{block}{}
    \justifying
    \textbf{Prop.} Let~$X$ and~$Y$ be well quasiorders.
    \good{Without} assuming~\bad{\textsc{lem}},~$X \times Y$ is well.
  \end{block}
  \pause

  \dnote{We have~$\mathsf{Well}(X,{\leq}) \to
  \mathsf{Well}_\infty(X,{\leq})$, but the converse only holds in presence of \hil{bar
  induction}. \emph{How much stronger exactly is the inductive rephrasing?}}
\end{frame}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
\begin{frame}{Topos for reifying generic models}
  Grothendieck toposes (= categories of sheaves over sites) are mathematical universes:
  \begin{itemize}
    \item The \hil{generic sequence} is an honest function~$\NN \to X$
    \hil{in} a certain topos~$\mathcal{E}$.
    \item The \hil{generic surjection} is an honest surjection \hil{in} an
    appropriate subtopos of~$\mathcal{E}$.
    \item In the \emph{double-negation subtopos} of the base, a predicate~$Q : \NN
    \to \mathsf{Prop}$ such that
    \[ Q\,x \times Q\,y \,\longrightarrow\, x = y \quad\text{and}\quad
      \neg\neg \sum_{n\?\NN} Q\,n \]
    looks like an ordinary natural number.
  \end{itemize}

  A statement holds in a topos iff a certain topos-directed translation holds
  in the base. For instance, for the double-negation subtopos the translation
  substitutes
  \[\begin{array}{rcl@{\quad\text{where}\quad}rcl}
    \exists & \leadsto & \exists^\mathrm{cl},
    & \exists^\mathrm{cl} &\defeqv& \neg\neg\exists, \\
    \vee & \leadsto & \vee^\mathrm{cl},
    & \alpha \vee^\mathrm{cl} \beta &\defeqv& \neg\neg(\alpha \vee \beta), \\
    = & \leadsto & =^\mathrm{cl},
    & s =^\mathrm{cl} t &\defeqv& \neg\neg(s = t).
  \end{array} \]
  In general, stage-dependent~$\nabla$ instead of~$\neg\neg$.
\end{frame}}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
\begin{frame}{Maximal ideals as convenient fictions}
  \jnote{1}{
    The theorem on the slide is a generalization of a fact from undergraduate
    linear algebra: Over a field, no surjective matrix can have more rows than
    columns. (``Surjective'' here means that the induced linear map is
    surjective.)
  }

  \jnote{6-7}{
    The slide presents a standard proof as offered by most textbooks on
    commutative algebra. The proof is quite efficient from a viewpoint of
    mathematical organization, as it quickly succeeds in reducing to the field
    situation. As such, it is short and memorable.

    However, the proof can also be critized for appealing to the transfinite
    two times; the methods of the proof are at odds with the concreteness of
    the statement of the theorem---from given equations witnessing
    surjectivity,~$Mv_i = e_i$, we are asked to deduce the equation~$1 = 0$.

    For this reason, the theorem and its classical proof are often used as case
    studies for tools and techniques aiming to extract constructive content
    from classical proofs. One such technique employs constructive forcing.
    The first(?)\@ constructive proof, found directly without using extraction
    techniques, is laid out in a
    \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{beautiful short note by Richman}.
  }

  \jnote{8-9}{
    For the countable case, there is an iterative construction of a maximal
    ideal, making do without any decidability assumptions (due to Krivine,
    later later clarified by Berardi and Valentini). It is a parlor trick,
    resulting in a subset which formally verifies the axioms for a maximal
    ideal but without carrying out any actual work. Indeed, the resulting ideal
    will in general not be a detachable subset of the ring.

    Surprisingly, there is still computational content in this construction, as
    explored in \fixedhref{https://arxiv.org/abs/2207.03873}{this joint paper
    with Peter Schuster}; one interpretation of our observation is that classical
    proofs don't ``really'' require a maximal ideal; they just use that notion
    for structuring hidden computations.
  }

  \jnote{9}{
    In a suitable topos, the ring appears countable. Hence we can
    carry out the iterative maximal ideal construction there. The resulting ideal
    will not be part of the base universe (instead, from the point of view of
    the base universe we will just have constructed a certain sheaf of ideals on a
    certain pointfree space), but bounded first-order consequences of its
    existence still pass down to the base.
  }

  \jnote{10}{
    Unwinding all the definitions from toposes and from the
    iterative maximal ideal construction, and eliminating the application
    of~\textsc{lem} from the classical proof presented before, we mechanically
    arrive at the constructive direct proof presented on the slide.
  }

  \begin{block}{}
    \textbf{Thm.}
    Let~$M$ be a surjective matrix with more rows than columns over a
    commutative ring~$A$. Then~$1 = 0$ in~$A$.
  \end{block}

  \only<1-9>{
    \visible<2->{\emph{Proof.} (classical) \bad{Assume not.}}
    \visible<3->{Then there is~a \bad{maximal ideal} $\mmm$.}
    \visible<5->{The matrix is surjective over~$A/\mmm$.}
    \visible<6->{Since~$A/\mmm$ is a field, this is a contradiction to basic linear algebra.\qed}
  }

  \only<4-9>{\medskip\par\centering\scalebox{0.9}{\centering\begin{tikzpicture}
    \node (0) at (0,1) {$(0) = \{0\}$};
    \node (1) at (0,5) {$(1) = \ZZ$};
    \node (2) at (-2,4) {$(2)$};
    \node [right of=2] (3) {$(3)$};
    \node [below of=2] (4) {$(4)$};
    \node [below of=2, xshift=0.7cm] (6) {$(6)$};
    \node [right of=3] (5) {$(5)$};
    \node [right of=5] (7) {$(7)$};
    \node [right of=7] (7d) {$\ldots$\phantom{(}};
    \node [right of=7d, xshift=3cm, yshift=-2cm] (max)
    {\vbox{\small{\it maximal among the proper ideals} \\ \medskip \hspace*{-6.75em}\textbullet \quad $\neg(1 \in
    \mmm)$ \\ \medskip \textbullet \quad $\neg\bigl(1 \in \mmm + (x)\bigr) \Rightarrow x \in \mmm$}};
    \node [below of=4] (8) {$(8)$};
    \node [right of=8, xshift=3cm] (8d) {$\ldots$};
    \draw (0) -- (8);
    \draw (0) -- (8d);
    \draw (0) -- (6);
    \draw (2) -- (1);
    \draw (3) -- (1);
    \draw (5) -- (1);
    \draw (7) -- (1);
    \draw (7d) -- (1);
    \draw (4) -- (2);
    \draw (8) -- (4);
    \draw (6) -- (2);
    \draw (6) -- (3);
    \draw [mypurple!30, thick, shorten <=-2pt, shorten >=-2pt, ->] (max) to [out=120, in=-30] (7d);
    \begin{pgfonlayer}{background}
      \draw[decorate, very thick, draw=mypurple!30]
        ($(2.south west) + (8pt, 0)$) arc(270:180:8pt) --
        ($(2.north west) + (0, -8pt)$) arc(180:90:8pt) --
        ($(7d.north east) + (-8pt, 0)$) arc(90:0:8pt) --
        ($(7d.south east) + (0, 8pt)$) arc(0:-90:8pt) --
        cycle;
    \end{pgfonlayer}
  \end{tikzpicture}\par}\par}

  \pause
  \pause
  \pause
  \pause
  \pause
  \pause
  \raggedright

  \only<7-9>{
    \emph{Does there exist a maximal ideal?}
    \pause
    \good{Yes}, if~$A$ is countable.
    In the general case: \bad{No}\pause, but \good{yes} in a \emph{suitable topos}, and
    \emph{bounded first-order consequences} of its existence there \good{pass down} to
    the base.
    \pause
  }

  \only<10>{
    \emph{Proof.} (constructive, special case) Write~$M =
    \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
    we have~$u, v \? A$ with
    \[
      u \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}1\\0\end{smallmatrix}\right)
      \quad\text{and}\quad
      v \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}0\\1\end{smallmatrix}\right).
    \]
    Hence
    $
      1 = (vy) (ux) = (uy) (vx) = 0
    $. \qed
  }
\end{frame}}

>>>>

\documentclass[12pt,utf8,notheorems,compress,t,aspectratio=169]{beamer}
\usepackage{etex}

@@

\usepackage{mathtools}
\usepackage{xspace}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{manfnt}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{proof}
\usepackage{agda}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning,svg.path}
\hypersetup{colorlinks=true}

@@

\title{Modal operators for a constructive account of well quasi-orders}

\author{Ingo Blechschmidt}
\date{November 30th, 2024}

@@

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\BB}{\mathbb{B}}
\newcommand{\pp}{\mathbbm{p}}
\newcommand{\MM}{\mathbb{M}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\FF}{\mathbb{F}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\CC}{\mathbb{C}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\ccc}{\mathfrak{c}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Cov}{\mathrm{Cov}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\inv}{inv.\@}
\newcommand{\forces}{\vDash}
\newcommand{\ind}{\ensuremath{_\text{ind}}\xspace}
\newcommand{\sinf}{\ensuremath{_\infty}}
\newcommand{\impl}{\ensuremath{_\text{impl}}\xspace}

\setbeamertemplate{blocks}[shadow=false]

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\includegraphics[width=\paperwidth]{tower}\end{minipage}}
\begin{frame}[c]
  \centering
  \color{white}

  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip

  \scriptsize

  \setbeamercolor{block body}{bg=black!100}
  \begin{minipage}{0.62\textwidth}
    \begin{block}{\centering Arbeitstagung Bern--München--Verona}
      \centering\normalsize\color{white}
      \hil{Modal operators for \\ a constructive account of
      well quasi-orders} \\[-0.9em]
      \
    \end{block}
  \end{minipage}

  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bf
  \colorbox{black}{November 30th, 2024}%
  \bigskip

  \colorbox{black}{\begin{minipage}{0.2\textwidth}
    \centering
    Ingo Blechschmidt \\
    University of Antwerp
  \end{minipage}}
\end{frame}}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{2.1em}\includegraphics[height=\paperheight]{sea-of-clouds-1}\end{minipage}}
\begin{frame}{Well quasi-orders}
  \textbf{Def.} Let~$(X,{\leq})$ be a quasi-order.
  \vspace*{-0.3em}
  \begin{itemize}
    \item A sequence~$\alpha : \NN \to X$ is \hil{good} iff
    there exist~$i < j$ with~$\alpha\,i \leq \alpha\,j$.
    \item The quasi-order~$X$ is \hil{well\only<4->{\sinf}} iff every sequence~$\NN \to X$ is good.
  \end{itemize}
  \pause
  \vspace*{-0.9em}

  \begin{columns}[t]
    \begin{column}{0.46\textwidth}
      \begin{block}{Natural numbers\phantom{y}}
        \textbf{Prop.} $(\NN, {\leq})$ is well\only<4->{\sinf}.
        \smallskip

        \emph{Proof.} Let~$\alpha : \NN \to \NN$. By~\badbox{\textsc{lem}}, there is a
        \bad{minimum}~$\alpha\,i$.
        Set~$j \defeq i + 1$. \qed

        \hfill\color{gray}offensive?
      \end{block}
    \end{column}
    \pause

    \begin{column}{0.50\textwidth}
      \begin{block}{Key stability results}
        \justifying
        \only<1-5>{Assuming~\badbox{\textsc{lem}}
        and~\badbox{\textsc{dc}}}\only<6->{Constructively}, \ldots

        \hil{Dickson:} \tabto{1.8cm} If~$X$ and~$Y$ are well\only<4-5>{\sinf}\only<6->{\ind}, so is~$X \times Y$. \\
        \hil{Higman:}  \tabto{1.8cm} If~$X$ is well\only<4-5>{\sinf}\only<6->{\ind}, so is~$X^\star$. \\
        \hil{Kruskal:} \tabto{1.8cm} If~$X$ is well\only<4-5>{\sinf}\only<6->{\ind}, so is~$\mathrm{Tree}(X)$.
      \end{block}
    \end{column}
  \end{columns}
  \pause
  \pause

  \textbf{Def.} A quasi-order~$X$ is \hil{well\ind} iff~$\mathsf{Good} \mid
  [\,]$, where~$\mathsf{Good}\,x_0 \ldots x_{n-1}$ iff~$\exists i < j\_ x_i \leq x_j$.
  \pause
  \pause

  \small
  With \bad{bar induction},\tabto{3.10cm} $\text{well}\ind \Leftarrow \text{well}\sinf$. \\[-0.2em]
  Constructively,\tabto{3.10cm} $\text{well}\ind \Rightarrow \text{well}\sinf$.
  \only<8>{
    \bigskip
    \begin{columns}[c]
      \begin{column}{0.01\textwidth}
	\includegraphics[height=2.4em]{question-mark}
      \end{column}
      \begin{column}{0.9\textwidth}
      Is there a procedure for reinterpreting \hil{classical proofs}
      regarding well\sinf{} as \newline \hil{blueprints for constructive proofs} regarding well\ind?
      \end{column}
    \end{columns}
  }
  \pause
  \pause
  Moreover, if~$X$ is well\ind, then \ldots
  \begin{itemize}
  \vspace*{-0.7em}
    \small
    \item for every \emph{partial} function~$\alpha$,\tabto{5.25cm}
    if~$\forall n\_ \neg\neg(\alpha\,n\,{\downarrow})$,
    then~$\neg\neg \exists i < j\_ \alpha\,i\,{\downarrow} \wedge
    \alpha\,j\,{\downarrow} \wedge \alpha\,i \leq \alpha\,j$. \\[-2.0em]\
    \item for every \emph{multivalued} function~$\alpha$,\tabto{5.25cm}
    $\exists i < j\_ \exists x \in \alpha\,i\_
    \exists y \in \alpha\,j\_ x \leq y$.
  \end{itemize}
  \pause

  \vspace*{-0.7em}
  \hil{Central insight:} A quasi-order $X$ is well\ind iff $\hil{$\necessary$}\,\forall\alpha : \NN \to
  X\_ \exists i < j\_ \alpha\,i \leq \alpha\,j$.
\end{frame}}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[width=\paperwidth]{staircase-more-faded}\end{minipage}}
\begin{frame}{The modal multiverse of constructive forcing}
  \textbf{Def.} A statement~$\varphi$ holds \ldots
  \begin{itemize}
    \small
    \item \hil{everywhere} \tabto{2.15cm}($\necessary\varphi$)\tabto{3.08cm} iff it holds in every topos
    (over the current base).
    \item \hil{somewhere} \tabto{2.15cm}($\possible\varphi$)\tabto{3.08cm} iff it holds in some positive topos.
    \item \hil{proximally} \tabto{2.15cm}($\xpossible\varphi$)\tabto{3.08cm} iff it holds in some positive overt topos.
  \end{itemize}
  \bigskip

  \begin{tikzpicture}[overlay]
    \node[anchor=south east,inner sep=0] (image) at (14.8,0.6) {
      \includegraphics[width=0.2\textwidth]{branching}
    };
  \end{tikzpicture}\vspace*{-2.3em}
  \pause

  \colorbox{white}{\bignumber{1} A quasiorder is well\ind iff \emph{everywhere}, every sequence is good.}

  \mbox{\qquad
  \colorbox{white}{\bignumber{2} A relation is well-founded iff \emph{everywhere}, there is
  no infinite descending chain.}}

  \qquad
  \qquad
  \colorbox{white}{\bignumber{3} A ring element is nilpotent iff all prime ideals \emph{everywhere} contain it.}

  \mbox{\qquad
  \qquad
  \qquad
  \colorbox{white}{\bignumber{4} For every inhabited set~$X$, \emph{proximally} there is an
  enumeration~$\NN \twoheadrightarrow X$.}}

  \qquad
  \qquad
  \qquad
  \qquad
  \colorbox{white}{\bignumber{5} For every ring, \emph{proximally} there is a maximal ideal.}

  \qquad
  \qquad
  \qquad
  \qquad
  \qquad
  \colorbox{white}{\bignumber{6} \emph{Somewhere}, the law of excluded middle holds.}
\end{frame}}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-4.68em}\includegraphics[width=1.2\paperwidth]{wqo-faded}\end{minipage}}
\begin{frame}{Answering a question by Berardi--Buriola--Schuster}
  %Stefano Berardi, Gabriele Buriola and Peter Schuster established that the
  %following notion is useful for structuring constructive arguments:

  \textbf{Def.} A quasi-order~$X$ is \hil{well\impl} iff (approximately) for
  every monotone predicate~$B$,
  \[ \text{if $B \mid_{\text{incr}} [\,]$, then $B \mid [\,]$.} \]
  \justifying
  {\small``Assume that no matter how the empty list evolves to an \emph{increasing}
  list~$\sigma$, eventually~$B\,\sigma$. Then no matter how the empty
  list evolves to an \emph{arbitrary} list~$\tau$, eventually~$B\,\tau$.''\par}
  \pause

  \hil{Equivalently:} If everywhere every \emph{increasing} infinite sequence~$\alpha : \NN
  \to X$ has a finite prefix validating~$B$, then so does every
  \emph{arbitrary} infinite sequence everywhere.
  \pause

  \textbf{Prop.} $\text{well\impl} \Rightarrow \text{well\ind}$.

  \emph{Proof.} Trivially~$\mathsf{Good} \mid_{\text{incr}} [\,]$,
  hence~$\mathsf{Good} \mid [\,]$ by assumption. \qed
  \pause

  \textbf{Thm.} $\text{well\impl} \Leftarrow \text{well\ind}$.

  \small
  \emph{Proof.} Let~$x_0,x_1,\ldots$ be an infinite sequence (in an arbitrary topos).
  \emph{Somewhere,} \textsc{lem} holds and implies that there is an increasing infinite subsequence~$x_{i_0},
  x_{i_1}, \ldots$. \emph{There} we have, by assumption, a finite prefix
  with~$B\,x_{i_0}\ldots x_{i_n}$. As~$B$ is monotone, \emph{there}
  we also have~$B\,x_0x_1\ldots x_{i_n}$. So \emph{somewhere} there is a finite
  prefix validating~$B$. Hence there actually is a finite prefix
  validating~$B$. \qed
\end{frame}}

\backupstart
{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-2.2em}\includegraphics[width=\paperwidth]{agda-wqo}\end{minipage}}
\begin{frame}
  \vspace*{\paperheight}\vspace*{-2.5em}\centering
  \emph{Agda formalization in progress.}
\end{frame}}
\backupend
